---
title: "Analyzing RNA-seq data with the DESeq2 package"
author: "Michael I. Love, Simon Anders, and Wolfgang Huber"
date: "`r doc_date()`"
package: "`r pkg_ver('DESeq2')`"
abstract: >
  A basic task in the analysis of count data from RNA-seq is the
  detection of differentially expressed genes. The count data are
  presented as a table which reports, for each sample, the number of
  sequence fragments that have been assigned to each gene. Analogous
  data also arise for other assay types, including comparative ChIP-Seq,
  HiC, shRNA screening, mass spectrometry.  An important analysis
  question is the quantification and statistical inference of systematic
  changes between conditions, as compared to within-condition
  variability. The package DESeq2 provides methods to test for
  differential expression by use of negative binomial generalized linear
  models; the estimates of dispersion and logarithmic fold changes
  incorporate data-driven prior distributions This vignette explains the
  use of the package and demonstrates typical workflows.
  [An RNA-seq workflow](http://www.bioconductor.org/help/workflows/rnaseqGene/)
  on the Bioconductor website covers similar material to this vignette
  but at a slower pace, including the generation of count matrices from
  FASTQ files.
output:
  BiocStyle::html_document2:
    toc: true
    toc_float: true
bibliography: library.bib
vignette: >
  %\VignetteIndexEntry{Analyzing RNA-seq data with the DESeq2 package (HTML version)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding[utf8]{inputenc}
---

```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(tidy=FALSE, cache=TRUE,
                      dev="png", dpi=180,
                      message=FALSE, error=FALSE, warning=TRUE)
```	

# Vignette porting note

<span style="color:red">This is an unfinished port of the package's
"old" Sweave/PDF vignette into Rmarkdown/HTML. Please consult the
[existing PDF vignette](DESeq2.pdf) in the mean time.</span>

# Standard workflow

**If you use DESeq2 in published research, please cite:**

> Love, M.I., Huber, W., Anders, S.,
> Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2, 
> *Genome Biology* 2014, **15**:550.
> [10.1186/s13059-014-0550-8](http://dx.doi.org/10.1186/s13059-014-0550-8)

Other Bioconductor packages with similar aims are
`r Biocpkg("edgeR")`,
`r Biocpkg("limma")`,
`r Biocpkg("DSS")`,
`r Biocpkg("EBSeq")`, and 
`r Biocpkg("baySeq")`.

## Quick start

Here we show the most basic steps for a differential expression analysis.
These steps assume you have a *RangedSummarizedExperiment* object
`se` which contains the counts and information about samples.
The `design` indicates that we want to
measure the effect of condition, controlling for batch differences.
The two factor variables `batch` and `condition` 
should be columns of `colData(se)`.

```{r quickStart, eval=FALSE}
dds <- DESeqDataSet(se, design= ~ batch + condition)
dds <- DESeq(dds)
res <- results(dds, contrast=c("condition","trt","con"))
```

* If you have a count matrix, the first line would use
  *DESeqDataSetFromMatrix* as shown below.
* If you have transcript quantification files, as produced by
  *Salmon*, *Sailfish*, or *kallisto*, you would use
  *DESeqDataSetFromTximport*, as shown below.

## How to get help

All DESeq2 questions should be posted to the Bioconductor support
site: <https://support.bioconductor.org>, which serves as a
repository of questions and answers. See the first question in the
list of Frequently Asked Questions (FAQ)
for more information about how to construct an informative post.

## Input data

### Why un-normalized counts?

As input, the DESeq2 package expects count data as obtained, e.g.,
from RNA-seq or another high-throughput sequencing experiment, in the form of a
matrix of integer values. The value in the *i*-th row and the *j*-th column of
the matrix tells how many reads can be assigned to gene *i* in sample *j*.
Analogously, for other types of assays, the rows of the matrix might correspond
e.g. to binding regions (with ChIP-Seq) or peptide sequences (with
quantitative mass spectrometry). We will list method for obtaining count matrices
in sections below.

The values in the matrix should be un-normalized counts of sequencing reads (for
single-end RNA-seq) or fragments (for paired-end RNA-seq). 
The [RNA-seq workflow](http://www.bioconductor.org/help/workflows/rnaseqGene/)
describes multiple techniques for preparing such count matrices.  It
is important to provide count matrices as input for DESeq2's
statistical model [@Love2014] to hold, as only the count values allow
assessing the measurement precision correctly. The DESeq2 model
internally corrects for library size, so transformed or normalized
values such as counts scaled by library size should not be used as
input.

### SummarizedExperiment input

The class used by the DESeq2 package to store the read counts 
is *DESeqDataSet* which extends the *RangedSummarizedExperiment*
class of the `r Biocpkg("SummarizedExperiment")` package.
The "Ranged" refer to the fact that the rows of the assay data (here
counts) are associated with genomic ranges (the exons of genes).
This facilitates downstream exploration of results, making use of
other Bioconductor packages' range-based functionality
(e.g. find the closest ChIP-seq peaks to the differentially expressed genes).

An example of the steps to produce a *RangedSummarizedExperiment* can
be found in the [RNA-seq workflow](http://www.bioconductor.org/help/workflows/rnaseqGene/)
and in the vignette for the data package `r Biocexptpkg("airway")`.
Here we load the *RangedSummarizedExperiment* from that package in
order to build a *DESeqDataSet*.

```{r loadSumExp}
library("airway")
data("airway")
se <- airway
```

A *DESeqDataSet* object must have an associated design formula.
The design formula expresses the variables which will be
used in modeling. The formula should be a tilde (~) followed by the
variables with plus signs between them (it will be coerced into an
*formula* if it is not already).  An intercept is included,
representing the base mean of counts. The design can be changed later, 
however then all differential analysis steps should be repeated, 
as the design formula is used to estimate the dispersions and 
to estimate the log2 fold changes of the model. 
The constructor function below shows the generation of a
*DESeqDataSet* from a *RangedSummarizedExperiment* `se`.

*Note*: In order to benefit from the default settings of the
package, you should put the variable of interest at the end of the
formula and make sure the control level is the first level.

```{r sumExpInput}
library("DESeq2")
ddsSE <- DESeqDataSet(se, design = ~ cell + dex)
ddsSE
```

# Session info

```{r sessionInfo}
sessionInfo()
```
# References

